Parameter-less Population Pyramid

### 2014-01-03 Parameterless Population Pyramid/main.tex
punch pass 1

\maketitle
\begin{abstract}
%Maximum 200 Words
A major limiting factor to many evolutionary techniques is their need to be reconfigured
to each new problem.  While some previous methods been able to remove or reduce
the need for parameter tuning, many do so by trading peak efficiency for general applicability.
The Parameter-less Population Pyramid (P3) strikes a balance between continuous integration
of diversity with extremely exploitative elitist operators, allowing it to solve easy
problems quickly and hard problems eventually.  When compared with three state of the art
optimization techniques across seven different problem classes P3 always finds the optimum
at least a constant factor \emph{\textbf{faster}} than the rest.  Even when optimally
tuning the comparisons, P3 has a lower growth complexity when test problems come
from a random distribution.  
Beyond empirical results, we suggest avenues for theoretical analysis of P3's
expected runtime.  In general, while these are only the first results for P3,
we feel they are indicative wide applicability.
\end{abstract}

@@

\section{Introduction}
Providing a good solution in reasonable time to a large set of real world black box problems
is a fundamental goal of evolutionary computation.  Achieving this goal requires an
optimizer to make assumptions about the nature of the problem is is solving.
%TODO NFL Citation?
Often built into the design of a genetic algorithm (GA) are evolutionary parameters,
which allow the user to tune the optimizer's assumptions to a specific problem.  These
parameters can allow a GA to efficiently search a multitude of landscapes.  However,
to do so the user must first determine how to correctly tune the parameters, which is
a complex and time consuming process unto itself.

The problem of parameter configuration has given rise to a number of attempts at
parameter-less optimization.  While many parameters (when using reasonable settings) are unlikely to effect the order
of complexity of an algorithm, population size can.  More specifically, sufficiently small
population sizes can cause exponential search times, with arbitrarily large population sizes
similarly stalling search.  Even more critically, setting the population size correctly relies
on difficult to discern features of each problem. In an early example, a crossover based GA was devised
which used concurrent racing populations to remove all user specified parameters~\cite{harik:1999:parameterlessga}.
This method of population racing has also been paired with more advanced evolutionary system,
such as Hierarchical Bayesian Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that while not optimal,
this method ensured the GA would take no more than a logarithmic increase in evaluations.
Furthermore, doing so would ensure that these optimizers could be generally applicable
without problem specific tuning.  Experimental evidence supports this theory,
with some evidence that in general the bound is closer to a constant factor worse
than optimal.

In this work we shall introduce the Parameter-less Population Pyramid (P3).  This iterative
method for constructing a collection of populations requires no user parameters.
It leverages entropy based linkage detection to learn how to efficiently mix
solutions without problem specific information beyond the evaluation function.  Perhaps
most importantly, unlike other parameter-less techniques it experimentally appears to be at least
a constant factor \emph{\textbf{improvement}} over comparable optimally configured
optimization methods.

\section{The P3 Algorithm}
\label{sec-p3}
Before going into detail about how each aspect of P3 works, consider the example
evolution presented in Figure~\ref{fig-p3}.  At the start of optimization there
is no known problem information, so solution \#1 is randomly generated.  This
is then brought to a local optima using a hill climber (Section~\ref{sec-hillclimber})
creating solution \#2, which is added to the pyramid (Section~\ref{sec-pyramid}) level
$P_0$.  As there is no further exploitable information, \#3 is randomly generated
and optimized to create \#4, which is added to $P_0$.  \#2 and \#4 are used to
define a crossover (Section~\ref{sec-crossover}), which improves \#4 to create
\#5. As \#5 used information from $P_0$ it is added to the next level of the pyramid, $P_1$.
Again P3 lacks information to continue, so \#6 is randomly generated and improved into
\#7.  Unlike \#2 and \#4, \#7 is not added in this example because it happens to be identical to
a solution already in the pyramid.  Even though \#7 is not unique, it still crosses with
$P_0$, which does make a unique improvement \#8 that is added to $P_1$ and mixed with that
level to create \#9.  As before, the process creates \#10, \#11, and \#12 using
random generation, hill climbing, and crossover, respectively.  In this example \#13
is not added to $P_2$ because it not a strict improvement in fitness over \#12.  There
is not enough information in $P_2$ to improve \#13 further, so the process would begin
yet again with a random solution.

@@

\subsection{Hill Climber}
\label{sec-hillclimber}
The terminology used to describe different methods of making minor modifications
to a solution, evaluating the quality of those changes, and iteratively constructing
new solutions, is full of duplicated, ambiguously defined terms.  As such we shall
describe what is likely an existing algorithm and give it a descriptive name that
may not match all of its previous incarnations.

For our purposes, we use the First Improvement Hill Climber (FIHC), a method of exploiting
local information through a series of modify-evaluate-accept iterations, to start
from randomly initialized solutions and end with single change local optimal results.
FIHC uses the following algorithm.  Consider all gene loci $x_i$ in a random order.
For a given $x_i$ try all possible alternative values for the gene $x_i$, evaluating
solution quality with each.  If any alternative value is a strict fitness improvement
over the current value for $x_i$, replace $x_i$ with the improved value.  If no
change is made to the solution, stop, otherwise iterate.

@@

FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
methods like Steepest Ascent Hill Climber~\cite{bosman:2011:lsbbo, goldman:2012:ltga}
which require $O(N^2)$ as changes are only accepted after all neighbor solutions
are evaluated.  In general FIHC is designed to find optima using a minimum number
of evaluations.  Furthermore, because FIHC tests loci in a random order it does
not cause any positional bias to the search.

\subsection{Pyramid}
\label{sec-pyramid}
Unlike many EC techniques, P3 does not have a single population of solutions.
Instead to maintains a pyramid like structure, such that each level of the pyramid
has a population of solutions, with each level representing something akin to different
generations of evolution.  More specifically, a solution's level is one higher than
its highest parent, similar to the layering done in~\cite{hornby:2006:alps}.

@@

\subsection{Crossover}
\label{sec-crossover}
P3 uses a crossover method derived from that defined for LTGA~\cite{thierens:2013:ltgahiff}
(Section~\ref{sec-ltga}),
in that it uses gene value entropy to construct clusters of genes that should have their values
conserved during crossover.  There are two primary steps in
defining this type of crossover: cluster creation and cluster usage.

During cluster creation, P3 attempts to create a binary tree of clusters such that the leaves
represent individual gene loci and each internal node represent the subset of loci created
by joining the subsets encoded in its children.  These clusters represent detected linked
loci in the genome, and are used when performing crossover.
Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
efficiently estimates the linkage between clusters of loci using the average pairwise
entropy, as given in Equation~\ref{eq-entropy}.
\begin{equation}
  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j} 
  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
  \label{eq-distance}
\end{equation}
\begin{equation}
  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
  \label{eq-entropy}
\end{equation}

Starting from the leaves, such that each loci appears exactly once, the binary tree
can be constructed using an agglomerative method.  Each iteration determine the
pairwise linkage between each cluster of loci using Equation~\ref{eq-distance},
and merge the two closest into a new cluster.  The process ends when there is only
a single cluster containing all loci.  As our clustering algorithm is both convex and
satisfies the swapping lemma~\cite{gronau:2007:upgma},
this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table.

Not all of the clusters in the linkage tree are worth using during crossover.  The most
obvious is a cluster containing all the loci.  If we performed a crossover such that all
of the loci must remain together, no new individual would be created.  We further remove
clusters that are completely contained within some other cluster, such that if
$D(C_i,C_j) = 0$, $C_i$ and $C_j$ are not used when performing crossover.  $D(C_i,C_j)$ only
returns $0$ when the loci in $C_i$ perfectly predict the bits in $C_j$ and vice versa,
and from this we conclude that the settings for those two loci should never be disrupted.
When used on particularly small populations this has the potential to greatly reduce the number
of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
at most two clusters exist: one for all genes where the solutions agree and one for where they disagree.

Given the useful set of clusters, we can now define how they are used.  P3 uses
crossover to recombine known information in a level of the pyramid with a candidate
solution.  For each cluster, do the following.  Select a random donor from the population.
Copy the donor's gene values for all loci in the cluster into the candidate solution.
Evaluate the candidate.  If the fitness of the candidate has been reduced, revert the candidate's
gene values.  To clarify, changes are kept in the case when the fitness is unaffected.
This allows the crossover to drift through fitness plateaus.

When choosing the donor P3 ensures there is at least one gene value
different between the donor and the candidate solution.  This can potentially require
up to $O(|C_j||P_i|)$, but empirically appears more efficient than allowing donations that
do not change any gene values.

To complete a single crossover event, all of the useful clusters are tested, each
requiring at most a single evaluation.  Clusters are tested in smallest first order,
with clusters of equal size ordered randomly.  This ensures smaller building blocks
are optimized before large changes are made to the genome, which helps preserve
diversity.  This also helps prevent genetic hitchhikers, which make interfere with
entropy calculations.  As cluster building and application ordering are not dependent on
the position of loci in the genome, there is no positional bias in the search points
evaluated.

\subsection{All Together}
\label{sec-alltogether}
The pieces of P3 were specifically chosen to interact in desirable ways.  First,
the hill climber provides an efficient tool for discovering pairwise linkage, as
search only ends when no single bit change can be made which improves fitness.
Therefore crossover, which relies on pairwise linkage, has high quality information
to leverage when determining clusters.

@@

Finally, the use of levels allows P3 to perform solution filtering without throwing
away information.  Once found, a local optima is stored indefinitely and its linkage
and genes will have impact on future crossovers.  However, by splitting solutions
into levels, the linkage information and the donated genes come from similar solution
qualities.  This prevents noise in gene linkage which only exists in low quality
solutions from effecting cluster formation for high quality solutions.  It also
reduces the chance for high quality solutions to simply overwrite new, unoptimized
solutions, preserving search diversity.

\begin{comment}
A major advantage to the design of P3 is that it makes very weak assumptions about
the search landscape.  In many evolutionary systems the crossover operator assumes
that genes linked in the search landscape are collocated in the genome.  This means
a problem changes difficulty depending on how the genome is set up, meaning the user
has to know information about gene linkage when applying the algorithm.  All of P3's
operations ignore gene ordering, removing that level of configuration from the user.
\end{comment}

\subsection{Contrasted With The Generational Model}
A major advantage of P3's structure is that it avoids the need to set a fixed
population size.  In a generational model, the diversity contained in the initial
population represents a bottleneck for future generations.  This is true
because in most systems, especially on deceptive problems, mutation rates are
low enough as to make the creation of large, multi-gene substructures unlikely
after initialization.  As such the generational model often results in a race between
decreasing diversity and increasing fitness, with the hope that the global optimum
will be found before too much diversity is lost.  As such, the population size
must be set large enough to contain sufficient initial diversity and
to maintain that diversity during search for just long enough for all of it to be
leveraged.

Premature convergence of a population leads to the generational model often having
two termination conditions: out of time or out of diversity.  In the former, if
a smaller population size had been used, more of the diversity could have been exploited
meaning a better solution likely could have been reached in the same number of evaluations.
In the latter, if a larger population size had been used more evaluations could have been
performed increasing the likelihood of improving solution quality.  Often the only option
is to restart, throwing away all of the previous search information.  In either case, unless
you can set the population size correctly beforehand, the generational model will likely
result in waste.

As P3 has exploited all known information each time the top of the pyramid is reached,
there is no fear of wasted evaluations on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence.

In general, P3 only makes two assumptions: local structure and non uniform gene interdependence.
In order for P3 to be efficient there needs to be at least some correlation between a
solution's fitness and the fitness of its single bit neighborhood.  This is required for
the hill climber to function properly.  Under this assumption, each time the hill climber
improves a solution it also improves the expected fitness of the neighborhood of the solution.
Stated another way, this assumes that small changes in genotype result in small changes in phenotype.
The non uniform gene interdependence assumption is required by the crossover operator to
perform clustering.  Without this assumption, the clusters formed by crossover will
all be due to spurious relations, and the donation of genes between solutions will be
no better than random modification.  Both assumptions are likely to hold true for
real world problems.  Also, many existing generational model systems have similar
or even more restrictive assumptions about search landscape.

\subsection{Order of Complexity}
With the exception of crossover, P3 requires no more than $O(N)$ computation
per evaluation, implying it is limited by the time required to evaluate a solution.
In order to perform crossover, P3 must spend $O(N^2)$ time updating pairwise
frequencies each time a solution is added to a level of the pyramid.  Similarly,
it requires $O(N^2)$ time to rebuild the set of crossover clusters once the frequency
table has been updated.  Na\"ively, this leads to the algorithm itself requiring $O(N^2)$ time.
However, multiple evaluations can occur for each time a single individual is added
to a population.

@@

To cause an update to $P_i$, all of the crossovers discovered for $P_{i-1}$ must
first be applied.  Each causes their own evaluation.  If P3 were to na\"ively use
all crossovers and perform an evaluation each time, this would cause $O(N)$ evaluations.
Again, this would result in $O(N)$ amortized cost per update to $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 ignore some clusters and will skip
an evaluation if no change to the solution was made by a crossover.  As such we currently
are unable to state conclusively that updating $P_i$ will require $O(N)$ evaluations.  That
said, empirically we expect it to be so.  Furthermore, it seems counterintuitive that
intentionally wasting evaluations would be considered beneficial to runtime.

@@

As P3 utilizes a hill climber to perform optimization, it is necessary to show P3's
added complexity is able to outperform using the hill climber alone.  Therefore
as the first comparison we define a complete optimization method based of off the
First Improvement Hill Climber defined in Section~\ref{sec-hillclimber}.  This definition
is extended to match P3's style of restarting. Whenever the hill climber
obtains a solution that cannot be improved using a single bit flip, a new solution
is randomly generated and optimized.

If P3's quality relies solely on starting at random points and applying hill climber,
this comparison algorithm should perform even better than P3.  Similar to P3 this
algorithm is Parameter-less, makes no assumptions about gene ordering, and will
find the global optima given a sufficient (potentially exponential) number of evaluations.

@@

LTGA works by iteratively improving a population of solutions in a generational manner.
Each generation half of the solutions (chosen using binary tournament selection)
are used to construct a linkage tree just as described in Section~\ref{sec-crossover}.
Each solution in the population is then crossed with the entire population using that
linkage tree, with the results put into the next generation.  This process continues
until the population fails to change between generations, signaling convergence.

Unlike P3, this variant of LTGA does not make use of a hill climber, and there is evidence that
doing so results in decreased efficiency~\cite{bosman:2011:lsbbo}.  LTGA applies
clusters in least linked first order whereas P3 uses smallest first.  Also, LTGA
does not search for donors until one containing different genes is found.  Instead
if the donated genes are identical, it simply skips the evaluation.  LTGA
requires a population size parameter, maintains a fixed population size across
generations, and does not prevent duplicate individuals in a population.  LTGA
does not add new genetic information after initialization, meaning that if
required diversity is not contained in the original population or lost in future
generations, LTGA will fail to find the global optimum.

In order to set the population size parameter to solve a class of problems, we use
the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
size required to meet a specified success criteria.  As P3 will run until the global optima
is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
on its success.  Therefore using the Rule of Three~\cite{jovanovic:1997:ruleofthree} we
can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
where $k$ is a number of runs without a failure.  Here we state that a population size
is successful if it performs 100 runs without failing to find the global optimum, meaning
the probability of that population failing to solve problems of the same class is bounded above by
$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
runs are not included during the comparative testing.  Furthermore, when testing on problem
classes, bisection is performed on a different set of instances than testing.

Other variants of LTGA have more closely resembled P3~\cite{goldman:2012:ltga},
but as those variants are not in active use, we did not use them for comparison.
However, no LTGA variant has been able to remove the population size parameter
or the problem of premature convergence.

@@

None of the methods have a bias toward creating specific gene values outside
of the bias created by the problem itself.  This means all of the optimizers
will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
an example, the problem of maximize the number of ones in a solution will be treated identically
to the problem of maximizing the number of zeros in a solution.  Both are also identical
to the problem of minimizing the hamming distance to any arbitrary solution.

>>>>

\maketitle
\begin{abstract}
%Maximum 200 Words
  A major limiting factor to practical application of evolutionary
  techniques is the tuning of their parameters which need to be
  reconfigured to each new problem.  While some previous methods been
  able to remove or reduce the need for parameter tuning, many do so
  by trading peak efficiency for general applicability.  The
  Parameter-less Population Pyramid (P3) is an evolutionary technique
  requiring \textbf{no} parameters and still broadly applicable to
  many problem types. It strikes a balance between continuous
  integration of diversity with exploitative elitist operators,
  allowing it to solve easy problems quickly and hard problems
  eventually.  When compared with three state of the art optimization
  techniques across seven different problem classes, P3 always finds
  the optimum at least a constant factor \emph{\textbf{faster}} than
  the rest.  Even when optimally tuning comparison technique
  parameters, P3 has a lower growth complexity when test problems come
  from a random distribution.  Beyond empirical results, we suggest
  avenues for theoretical analysis of P3's expected runtime.  In
  general, while these are only the first results for P3, we feel they
  are indicative wide applicability.
\end{abstract}

@@

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black box problems is a fundamental goal of evolutionary
computation.  Achieving this goal often requires an optimizer to make
assumptions about the nature of the problem is is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.

Often built into the design of a genetic algorithm (GA) are
evolutionary parameters which allow the user to tune the optimizer's
assumptions concerning a specific problem.  These parameters can allow
a GA to more efficiently search the problem's landscapes.  However, to
do so the user must first determine how to correctly \textbf{tune} the
parameters, typically a complex and time consuming process unto
itself.

Some considerable work has been devoted to understanding how to tune
these parameters. Early work in Evolutionary Strategies introduced the
concept of self-adaptive evolution which was subsequently introduced
to genetic algorithms~\cite{Back:1992:selfadapt}. Here, each solution
carries some of the parameters used and those parameters are adapted
along with the solution they are co-encoded with. Other work introduced the concept
of a \emph{meta} algorithm\cite{grefenstette:1986:optimalga}. Here a
second optimizer runs ``on top'' of the actual problem-solving
optimizer, the second slowly tuning the parameters of the first. Both
of these approaches are rather slow, in particular the meta
approach, as actual problem solving must occur at set levels to give
feedback as to how to change the parameters (online or to the meta
optimizer).

Many of these parameters have ``reasonable'' settings that are
unlikely to effect the order of complexity of an genetic algorithm,
but population size is an exception. For example, a population sized
too small can cause exponential search times, while arbitrarily large
population sizes can lead to wasted evaluations and stalled search.
More importantly, setting the population size correctly relies on
difficult to discern features of each problem. Some research has been
done on setting optimal population
sizes. Goldberg~\cite{goldberg:1991:gasize} showed that small
population sizes were not capable of finding good solutions, though
some practitioners~\cite{haupt:2000:optimum} have found that small
populations with a large mutation rate are effective on practical
problems.

All of these complications have lead to research on how to rid
evolutionary/genetic algorithms of parameter settings of any kind.  In
an early example, a crossover based GA was devised which used
concurrent racing populations to remove all user specified
parameters~\cite{harik:1999:parameterlessga}.  This method of
population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that
while not optimal, this method ensured the GA would take no more than
a logarithmic increase in evaluations.  Furthermore, doing so would
ensure that these optimizers could be generally applicable without
problem specific tuning.  Experimental evidence supports this theory,
with some evidence that in general the bound is closer to a constant
factor worse than optimal.

In this work we shall introduce the Parameter-less Population Pyramid
(P3).  P3 is an iterative method for constructing a \textbf{collection} of
populations that requires no user parameters.  It leverages entropy based
linkage detection to learn how to efficiently mix solutions without
problem specific information beyond the evaluation function.  Perhaps
most importantly, unlike other parameter-less techniques it
experimentally appears to be at least a constant factor
\emph{\textbf{improvement}} over comparable optimally configured
optimization methods.

\section{The P3 Algorithm}
\label{sec-p3}
Before going into detail about how each aspect of P3 works, let us
consider an example of P3 operation. Unlike many EC techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization there
are no solutions so solution \#1 is randomly generated.  Solution \#1
is brought to a local optima using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, little more can be done
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyrimad at population level $P_0$.  With a population of two,
crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5, which is an improvement. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it not a strict improvement in fitness over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

@@

\subsection{Hill Climber}
\label{sec-hillclimber}
% The terminology used to describe different methods of making minor modifications
% to a solution, evaluating the quality of those changes, and iteratively constructing
% new solutions, is full of duplicated, ambiguously defined terms.  As such we shall
% describe what is likely an existing algorithm and give it a descriptive name that
% may not match all of its previous incarnations.

For our purposes, we use a First Improvement Hill Climber
(FIHC)\footnote{We use the name FHIC for an algorithm that likely has
  various names in the literature, but use it here with a descriptive
  name for clarity.}, a method of exploiting local information through
a series of modify-evaluate-accept iterations starting from a randomly
initialized solution and ending with a optimum, either local or global.
FIHC uses the following algorithm.  Consider all gene loci
$x_i$ in a random order.  For a given $x_i$ try all possible
alternative values for the gene $x_i$, evaluating solution quality
with change.  If any alternative value is a strict fitness improvement
over the current value for $x_i$, replace $x_i$ with the improved
value.  Iteration of the process continues until no such change results in
a strict fitness improvement.

@@

FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
methods like Steepest Ascent Hill Climber~\cite{bosman:2011:lsbbo, goldman:2012:ltga}
which require $O(N^2)$ as changes are only accepted after all neighbor solutions
are evaluated.  In general, FIHC is designed to find optima using a minimum number
of evaluations.  Furthermore, because FIHC tests loci in a random order it does
not cause any positional bias to the search.

\subsection{Pyramid}
\label{sec-pyramid}
As mentioned, P3 maintains a pyramid like structure of populations
with each level representing something akin to different generations
of evolution. More specifically, a solution's level is one higher than
its highest parent, similar to the layering done
in~\cite{hornby:2006:alps}.

@@

\subsection{Crossover}
\label{sec-crossover}
P3 uses a crossover method derived from that defined for LTGA~\cite{thierens:2013:ltgahiff}
(Section~\ref{sec-ltga}),
P3 uses gene value entropy to construct clusters of genes that should have their values
conserved during crossover.  There are two primary steps in
defining this type of crossover: \emph{cluster creation} and \emph{cluster usage}.

During cluster creation, P3 attempts to create a binary tree of clusters such that the leaves
represent individual gene loci and each internal node represent the subset of loci created
by joining the subsets encoded in its children.  These clusters represent linked
loci that have been detected in the genome, and are used when performing crossover.
Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
efficiently estimates the linkage between clusters of loci using the average pairwise
entropy, as given in Equation~\ref{eq-entropy}.
\begin{equation}
  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j} 
  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
  \label{eq-distance}
\end{equation}
\begin{equation}
  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
  \label{eq-entropy}
\end{equation}

Starting from the leaves, the binary tree can be constructed using an
agglomerative method.  Each
iteration determines the pairwise linkage between each cluster of loci
using Equation~\ref{eq-distance}, and merge the two closest into a new
cluster.  The process ends when there is only a single cluster
containing all loci.  As our clustering algorithm is both convex and
satisfies the swapping lemma~\cite{gronau:2007:upgma}, this algorithm
requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table.

Not all of  the resulting clusters in the linkage  tree are useful for
crossover.  The most obvious useless  cluster is one that contains all
the loci.   A crossover that preserves  all loci in  a solution cannot
create  a  new  individual.   We  further  remove  clusters  that  are
completely  contained   within  some  other  cluster,   such  that  if
$D(C_i,C_j)  =  0$, $C_i$  and  $C_j$  are  not used  when  performing
crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
perfectly predict the  bits in $C_j$ and vice versa. From this we can
conclude  that  the  settings  for  those two  loci  should  never  be
disrupted.  When  used on particularly small populations  this has the
potential  to  greatly  reduce  the  number of  clusters  used  during
crossover.  For  instance, if  $|P_i| = 1$,  no clusters exist. If
$|P_i| =  2$ at most two clusters  exist: one for all  genes where the
solutions agree and one for where they disagree.

Given a useful set of clusters, we can now define how they are used.
P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution.  The algorithm is as follows. For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in the cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through plateaus.

When choosing a donor, P3 ensures there is at least one gene value
different between the donor and the candidate solution.  This can
potentially require up to $O(|C_j||P_i|)$, but empirically appears
more efficient than allowing donations that do not change any gene
values.

To complete a single crossover event, all of the useful clusters are tested, each
requiring at most a single evaluation.  Clusters are tested in smallest first order,
with clusters of equal size ordered randomly.  This ensures smaller building blocks
are optimized before large changes are made to the genome, thus helping preserve
diversity.  This also helps prevent genetic hitchhikers, which make interfere with
entropy calculations.  As cluster building and application ordering are not dependent on
the position of loci in the genome, there is no positional bias in the search points
evaluated.

\subsection{All Together}
\label{sec-alltogether}
The pieces of P3 were specifically chosen to provide important capabilities.  First,
the hill climber provides an efficient tool for discovering pairwise linkage, as
search only ends when no single bit change can be made which improves fitness.
Therefore crossover, which relies on pairwise linkage, has high quality information
to leverage when determining clusters.

@@

Finally, the use of levels allows P3 to perform solution filtering
without throwing away information.  Once found, a local optima is
stored indefinitely, preserving both its linkage and genes for future
crossovers.  By splitting solutions into levels, the linkage
information and the donated genes come from similar solution
qualities.  This prevents noise in gene linkage which only exists in
low quality solutions from effecting cluster formation for high
quality solutions.  It also reduces the chance for high quality
solutions to simply overwrite new, unoptimized solutions, preserving
search diversity.

  A major advantage to the design of P3 is that it makes very weak
  assumptions about the search landscape.  In many evolutionary
  systems the crossover operator assumes that genes linked in the
  search landscape are collocated in the genome.  This introduces an
  initialization problem, changing the difficulty of the problem due
  to a, potentially unknown, ordering required on the genome.  All of
  P3's operations ignore gene ordering, removing that level of
  configuration from the user.

\subsection{Contrasted With The Generational Model}
A major advantage of P3's structure is that it avoids the need to set
a fixed population size.  In a generational model, the diversity
contained in the initial population represents a bottleneck for future
generations.  This is true because in most systems, especially on
deceptive problems, mutation rates are low enough as to make the
creation of large, multi-gene substructures unlikely after
initialization.  As such the generational model often results in a
race between decreasing diversity and increasing fitness. As such, the
population size must be set large enough to contain sufficient initial
diversity and to maintain that diversity during search for just long
enough for all of it to be leveraged.

Premature convergence of a population leads to the generational model
often having two termination conditions: out of time or out of
diversity.  In the former, if a smaller population size had been used,
more of the diversity could have been exploited meaning a better
solution likely could have been reached in the same number of
evaluations.  In the latter, if a larger population size had been
used, more evaluations could have been performed increasing the
likelihood of improving solution quality.  In
either case, unless the population size is correctly set, the
generational model will likely result in waste.

As P3 has exploited all known information each time the top of the pyramid is reached,
there no wasted evaluations on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence. It is this essential balance of
evaluation to diverstiy that P3 focuses on.

In general, P3 only makes two assumptions: local structure and non uniform gene interdependence.
In order for P3 to be efficient there needs to be at least some correlation between a
solution's fitness and the fitness of its single bit neighborhood.  This is required for
the hill climber to function properly.  Under this assumption, each time the hill climber
improves a solution it also improves the expected fitness of the neighborhood of the solution.
Stated another way, this assumes that small changes in genotype result in small changes in phenotype.
The non uniform gene interdependence assumption is required by the crossover operator to
perform clustering.  Without this assumption, the clusters formed by crossover will
detect spurious relations. The resulting donation of genes between solutions will be
no better than random modification.  Both assumptions are likely to hold true for
real world problems and are similarly assumed by most generational models.

\subsection{Order of Complexity}
With the exception of crossover, P3 requires no more than $O(N)$ computations
per evaluation, implying it is limited by the time required to evaluate a solution.
In order to perform crossover, P3 must spend $O(N^2)$ time updating pairwise
frequencies each time a solution is added to a level of the pyramid.  Similarly,
it requires $O(N^2)$ time to rebuild the set of crossover clusters once the frequency
table has been updated.  Na\"ively, this leads to the algorithm itself requiring $O(N^2)$ time.
However, multiple evaluations can occur for each time a single individual is added
to a population.

@@

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied.  Each causes their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $O(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost per update to $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a crossover.  As such we currently are unable to state conclusively
that updating $P_i$ will require $O(N)$ evaluations.  That said,
empirically it to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

@@

As P3 utilizes a hill climber to perform optimization, it is necessary
to show P3's added complexity is able to outperform an optimizer that
uses a hill climber alone.  Therefore as the first comparison we
define a complete optimization method based on the First Improvement
Hill Climber defined in Section~\ref{sec-hillclimber}.  This
definition is extended to match P3's style of restarting. Whenever the
hill climber obtains a solution that cannot be improved using a single
bit flip, a new solution is randomly generated and optimized.

If P3's quality relies solely on starting at random points and
applying FIHC, this comparison algorithm should perform even better
than P3.  Like P3, FHIC-alone algorithm is Parameter-less, makes no
assumptions about gene ordering, and will find the global optima given
a sufficient (potentially exponential) number of evaluations.

@@

LTGA works by iteratively improving a population of solutions in a generational manner.
Each generation, half of the solutions (chosen using binary tournament selection)
are used to construct a linkage tree just as described in Section~\ref{sec-crossover}.
Each solution in the population is then crossed with the entire population using that
linkage tree, with the results put into the next generation.  This process continues
until the population fails to change between generations, signaling convergence.

Unlike P3, this variant of LTGA does not make use of a hill climber,
and there is evidence that doing so results in decreased
efficiency~\cite{bosman:2011:lsbbo}.  LTGA applies clusters in least
linked first order whereas P3 uses smallest first.  Also, LTGA does
not search for donors until one containing different genes is found.
Instead, if the donated genes are identical, it simply skips the
evaluation.  LTGA requires a population size parameter, maintains a
fixed population size across generations, and does not prevent
duplicate individuals in a population.  LTGA does not add new genetic
information after initialization, meaning that if required diversity
is not contained in the original population or lost in future
generations, LTGA will fail to find the global optimum.

Since LTGA requires a population size parameter, we thought it most fair to
compare against LTGA with an optimized population size, even though P3
requires no such tuning. To set the population size parameter, we use
the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
size required to meet a specified success criteria.  As P3 will run until the global optima
is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
on its success.  Therefore, using the Rule of Three~\cite{jovanovic:1997:ruleofthree}, we
can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
where $k$ is a number of runs without a failure.  Here we state that a population size
is successful if it performs 100 runs without failing to find the global optimum, meaning
the probability of that population failing to solve problems of the same class is bounded above by
$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
runs are not included during the comparative testing.  Furthermore, when testing on problem
classes, bisection is performed on a different set of instances than testing.

Other variants of LTGA have more closely resembled P3~\cite{goldman:2012:ltga},
but as those variants are not in active use, we did not use them for comparison.
However, no LTGA variant has removed the population size parameter
or fully addressed the problem of premature convergence.

@@

None of the methods have a bias toward creating specific gene values outside
of the bias created by the problem itself.  This means all of the optimizers
will have unchanged behavior if a problem's inputs are reflected or XOR'd with a fixed string.  As
an example, the problem of \emph{maximize the number of ones} in a solution will be treated identically
to the problem of \emph{maximize the number of zeros} in a solution.  Both are also identical
to the problem of minimizing the hamming distance to any arbitrary solution.
